Covariate analysis of MSSM reprocessed counts (with CQN normalisation and covariates adjusted)

### 2017-03-23 code/Rmd/MSSM_REPROCESSED.Rmd
Modified MSSM reprocessing to include modified algorithms for SVA and samples with picard reruns

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn7554308'
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of MSSM reprocessed RNASeq data';

@@

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Get reprocessed sample counts for temporal cortex
COUNT_ID <- 'syn7994853';
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F)

@@

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA = METADATA %>%
  dplyr::mutate(AOD = gsub('\\+','',AOD),
                BrainRegion = factor(BrodmannArea,
                                     levels = c('BM10', 'BM22', 'BM36', 'BM44'),
                                     labels = c('BM10' = 'FP', 'BM22' = 'STG', 
                                                'BM36' = 'PHG', 'BM44' = 'IFG')),
                Diagnosis = 'NONE', 
                RIN2 = RIN^2)
METADATA$Diagnosis[METADATA$CDR <= 1] = 'ND'
METADATA$Diagnosis[METADATA$CDR == 2 | METADATA$CDR == 3] = 'MD'
METADATA$Diagnosis[METADATA$CDR >= 4] = 'SD'
levels(METADATA$batch)[levels(METADATA$batch) == ''] = 'NoBatch'

@@

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering, cache=TRUE}
primaryVariable <- "BrainRegion.Diagnosis"
FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AOD","PMI", "PCT_PF_READS", "PCT_PF_READS_ALIGNED", 
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", 
                    "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
```{r preadj.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

@@

# Calculate the number of surrogate variables
NUM.SV = sva::num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                     MODEL1, method = 'be', seed = 123456, B = 30)

@@

  tissueTypeAbrv	= 'FP, STG, PHG, IFG', 
  study = 'MSSMRNASeq', 
  center = 'MSSM',

>>>>

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn7554308'
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of MSSM reprocessed RNASeq data';

@@

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data}
# Get reprocessed sample counts for temporal cortex
COUNT_ID <- 'syn7994853';
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F)

@@

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Choose samples with rRNA rate less than 5% and more PF_READS_ALIGNED
METADATA = METADATA %>%
  ddply(.(barcode), .fun = function(x){
    x %>%
      dplyr::filter(PCT_RIBOSOMAL_BASES < 0.05) %>%
      top_n(1, PF_READS_ALIGNED)
  })

# Fix metadata
METADATA = METADATA %>%
  dplyr::mutate(AOD = gsub('\\+','',AOD),
                BrainRegion = factor(BrodmannArea,
                                     levels = c('BM10', 'BM22', 'BM36', 'BM44'),
                                     labels = c('BM10' = 'FP', 'BM22' = 'STG', 
                                                'BM36' = 'PHG', 'BM44' = 'IFG')),
                Diagnosis = 'NONE', 
                RIN2 = RIN^2)
METADATA$Diagnosis[METADATA$CDR <= 1] = 'ND'
METADATA$Diagnosis[METADATA$CDR == 2 | METADATA$CDR == 3] = 'MD'
METADATA$Diagnosis[METADATA$CDR >= 4] = 'SD'
levels(METADATA$batch)[levels(METADATA$batch) == ''] = 'NoBatch'

@@

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering}
primaryVariable <- "BrainRegion.Diagnosis"
FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AOD","PMI",  "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
```{r preadj.covariates}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

@@

# Source modified functions from sva package
# source('../R/num.sv.R')
source('../R/irwsva.build.R')
source('../R/f.pvalue.R')

# Calculate the number of surrogate variables
# NUM.SV = num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
#                      MODEL1, method = 'be', seed = 123456, B = 30, tol = 1e-20)

# Get residuals from differential expression model
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, VOOM.WEIGHTS$E)

# Compute actual variance of all principal components
tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
actual.var = tmp$d^2/sum(tmp$d^2)

# Compute permuted variance of all principal components
permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
  tmp = svd(tmp.residual)
  permuted.var = tmp$d^2/sum(tmp$d^2)
}
permuted.var = apply(permuted.var, 2, mean)
NUM.SV = sum(actual.var >= permuted.var[1])

# Plot variance components
var.comp = data.frame(component = 1:length(actual.var), 
                      residual = actual.var, 
                      permuted.residual = permuted.var) %>%
  tidyr::gather(data, value, -component)
p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.2)), aes(x = component, y = value, color = data)) + geom_point()
p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
p

@@

  tissueTypeAbrv	= 'FP, STG, PHG, IFG', 
  study = 'MSSMRNASeq', 
  center = 'MSBB',
